Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-20 paper.tex
Update text for 1kx10k Zarr versions

\begin{figure}
\begin{center}
\includegraphics[]{figures/data-scaling}
\end{center}
\caption{Compression performance on simulated genotypes.
Comparison of total stored bytes for VCF data produced 
by subsets of a large simulation of French-Canadians.
Sizes for $10^6$ samples are shown on the right. Sizes 
for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
Also shown for reference is the size of genotype matrix 
when encoded as two bits per diploid genotype (2bit), as used 
in the PLINK binary format.
\label{fig-data-storage}}
\end{figure}

@@

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
almost identical compression performance in this example.
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

\begin{figure}[h]
\includegraphics{figures/whole-matrix-decode}
\caption{Genotype decoding performance.
Total CPU time required to decode genotypes into memory using the Zarr-Python
and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
Elapsed time is also reported (dotted line). 
This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
3.9 GiB/s Zarr (Zstd), 4.1 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
\label{fig-whole-matrix-decode}}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[]{figures/data-scaling}
\end{center}
\caption{Compression performance on simulated genotypes.
Comparison of total stored bytes for VCF data produced 
by subsets of a large simulation of French-Canadians.
Sizes for $10^6$ samples are shown on the right.
Also shown for reference is the size of genotype matrix 
when encoded as two bits per diploid genotype (2bit), as used 
in the PLINK binary format. Default compression settings were 
used in all cases.
\label{fig-data-storage}}
\end{figure}

@@

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
very similar compression performance in this example (23.75GiB
and 21.25GiB for $10^6$ samples, respectively). When using
a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
storage is reduced to 22.07GiB (not shown).
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

\begin{figure}[h]
\includegraphics{figures/whole-matrix-decode}
\caption{Genotype decoding performance.
Total CPU time required to decode genotypes into memory using the Zarr-Python
and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
Elapsed time is also reported (dotted line). 
This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
2.7 GiB/s Zarr (Zstd), 2.9 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
\label{fig-whole-matrix-decode}}
\end{figure}
